Engineering momentum profiles of cold-atom beams

### 2018-10-17 Paper/version2/paper.tex
Minor progress on methods section.

\usepackage{epsfig}
\usepackage{color}
\usepackage{graphicx}
\usepackage{bm}
\usepackage{epstopdf}

@@

\subsection{Finding a Link Potential}
Our simple process for finding an appropriate link potential consists of a global search over a family of possible link potentials in order to minimize the difference (more precisly the ``cost'', defined in Eq.~\eqref{eq:cost} below) between the desired and the actual flux properties. We perform this search using the well-known global optimization routine called Differential Evolution (DE) \cite{original DE paper}.This evolutionary-based search is suitable given the non-convex nature of the optimization problem. Despite its simplicity, DE does a good job of balancing the need to search the entire space against the need to learn from subsequent trials in the optimization process in order to speed up convergence with minimal complication in terms of choosing the right model settings. Emperically, we found DE to perform much better than several other approaches including random search, Nelder-Mead, and Simulated Annealing \cite{Tests performed using mathematica}.

Before detailing our procedure, we emphasize that our goal is not to the find the ``best'' link potential in terms of a theoretical cost function; one that is ``good enough'' in terms of experimental requirements will do. This relaxed goal removes many complications. For instance, as long as a discovered potential is {\it good enough} we do not even need to worry about whether or not the DE routine has converged. Similarly, if we run the routine twice and arrive at two very different but similarly {\it good enough} solutions, we do not have to worry about which solution is ``right''. We can just pick our favorite and proceed.

To start, we must specify the ideal desired flux profile $T(k)$ where $k$ is the incident momentum. This could be, for instance, a ``comb'' profile that only allows particle flux near a specified momentum, but our prescription applies to a broad class of flux profiles.

We also must specify the form of the link potential, $V(x, \bm{\theta})$, with unknown parameters $\bm{\theta}$. The functional form of the link potential should be chosen to represent the class of physically realizable potentials within the experimental context. In our case, we assume a sum of Gaussian potentials:
\begin{equation}
V(x,\{\bm{A},\bm{\mu},\bm{\sigma} \}) = \sum_{i=1}^{N}\frac{A_i}{\sqrt{2\pi\sigma_i^2}}\exp\left[{-\frac{(x-\mu_i)^2}{2\sigma_i^2}}\right],
\end{equation}
where $N$ is the number of Gaussian potentials (and the length of the parameter vectors $\bm{A}$, $\bm{\mu}$, and $\bm{\sigma}$).

In addition to specifying the form of the link potential, we may also need to specify a set of constraints on the parameters themselves in order to reflect the limitations in creating the physical potentials. In our case, we will enforce the following constraints
\begin{align}
  \sum_{i=1}^{N}\mu_i &= 0 & \\
  \sigma_i &> 0 &\forall i \in\{1,2,\ldots,N\}\\
  0 &\leq A_i < A^\mathrm{max} &\forall i \in\{1,2,\ldots,N\}             
\end{align}

In addition there may be generic constraints on the potential itself

@@

>>>>

\usepackage{epsfig}
\usepackage{color}
\usepackage{array}
\usepackage{graphicx}
\usepackage{bm}
\usepackage{epstopdf}

\DeclareMathOperator*{\argmin}{argmin}

@@

\subsection{Procedure for Finding a Link Potential}\label{sec:procedure_link}
We find an appropriate link potential $V_0(x) \equiv V(x,\bm{\theta}^*)$ by performing a global search over a family of possible potentials $V(x,\bm{\theta})$ for the parameters $\bm{\theta}^*$ that reduce the difference between the desired transmission-momentum profile $T_0(k)$ and the actual profile $T_{\bm\theta}(k)$ determined by a sample potential $V(x, \bm{\theta})$. Concretely, we minimize the cost $J_{\bm{\theta}}$ which is a function of $T_0(k) - T_{\bm{\theta}}(k)$ but may also explicitly depend upon $V(x,\bm{\theta})$ and $\bm{\theta}$ in order to encode practical, experimental constraints on the solutions for the link potential. Given the cost function, the {\it optimal} parameters are $\bm{\theta}^* = \argmin_{\bm{\theta}}J_{\bm\theta}$. However, we do not need to find the {\it optimal} link potential; one that is {\it good enough} in terms of experimental requirements will do. This relaxed goal alleviates concerns about convergence to the global optimum.

We perform the minimization outlined above using the global optimization routine called Differential Evolution (DE) \cite{original DE paper}. This evolutionary-based search algorithm is suitable given the non-convex (multiple local minima) nature of the optimization problem. Despite its simplicity, DE does a good job of balancing exploration of the space of link potentials against the need to efficiently learn from each sample with little tuning of the model settings. Emperically, we found DE to perform much better than several other approaches including random search, Nelder-Mead, and Simulated Annealing \cite{Tests performed using mathematica}.

We parametrize the family of link potentials $V(x, \bm{\theta})$, as a sum of $N$ Gaussian potentials:
\begin{equation}\label{eq:V-param}
V_N(x,\{\bm{A},\bm{\mu},\bm{\sigma} \}) = \sum_{i=1}^{N}\frac{A_i}{\sqrt{2\pi\sigma_i^2}}\exp\left[{-\frac{(x-\mu_i)^2}{2\sigma_i^2}}\right],
\end{equation}
where $A_j\equiv(\bm{A})_j$, $\mu_j\equiv(\bm{\mu})_j$, and $\sigma_j\equiv(\bm{\sigma})_j$ are the amplitude, mean, and standard deviation for the $j^{\mathrm{th}}$ Gaussian. It is necessary to enforce constraints on these parameters to both improve convergence and to ensure that solutions satisfy the experimental requirements. Table~\ref{tab:constraints} shows the explicit parameter constraints that we enforce  during minimization.

\begin{table}[t]
  \renewcommand*{\arraystretch}{1.4}
  \begin{tabular}{m{3cm}|m{5.5cm}}
    Constraints & Experimental Rationale \\
    \hline\hline
    $\sum_{i=1}^{N}\mu_i = 0$ & The cost function has a continuous degeneracy associated with overall translations of the link potential. \\
    \hline
    $\sigma_{\mathrm{min}} \leq \sigma_j \leq \sigma_{\mathrm{max}} $ & Laser beam widths fall between a minimum and maximum value.\\
    \hline
    $A_{\mathrm{min}} \leq |A_j| \leq A_{\mathrm{max}}$ & Laser amplitudes fall between a minimum and maximum value.
  \end{tabular}
  \caption{The explicit constraints on the potential parameters and the rationale for each constraint. The values of $\sigma_{\mathrm{min}}$, $\sigma_{\mathrm{max}}$, $A_{\mathrm{min}}$, and $A_{\mathrm{max}}$ must be determined from the experimental context.}
  \label{tab:constraints}
\end{table}

In addition to explicit constraints on the parameters, we inforce an implicit constraint by requireing that the link potential should not extend beyond the region  of potential support $x\in[-x_0,x_0]$. This we do by adding to the cost function a term $J_{\mathrm{boundary}}$ that grows with the area of the potential outside the support region. The complete definition of the cost function is given in suppl.~material.

@@

\subsection{Cost Function}\label{sec:cost_function}

\begin{equation}
  J'_{\mathbf{\theta}} = J_{\mathbf{\theta}} + \alpha J_{\mathrm{boundary}},
\end{equation}
where $\alpha>0$ is the weight given to cost incurred for having the potential extend outside the support region of the potential.
